Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-20 index.qmd
section:background,structural,dag; process:write

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T=\{1,2\})$. Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding. The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that maps data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness in comparing two proportions, yielding accurate estimates when its underlying assumptions are met [@Kanji_2006]. The Z-test is expressed as a signal-to-noise ratio: $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is the difference between the group sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y| T=1]$ and $E[Y| T=2]$, respectively. The noise $\hat{s}_{p}$ is the unpooled sample variability observed between the two groups.

@@

<!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
Experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. However, researchers often face constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these situations, causal inference provides a valuable alternative for generating causal estimates, particularly when the goal is to understand the mechanisms underlying specific data. Moreover, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].

<!-- 5. What is causal inference? -->
Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, *causal inference* is a framework designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. This framework employs rigorous mathematical techniques to address the fundamental problem of causality [@Pearl_2009], which is essential for understanding and defining causal effects. The *fundamental problem of causality* centers on answering the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals. *Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a specific cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].

<!-- 6. But how does it work? --> 
Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality is possible using the previous example. The framework starts by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$. Notice once a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and is thus termed a *counterfactual*. To address the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE). The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, and represents the difference between the average of observed potential outcomes and counterfactuals in the sample. Finally, akin to experimental studies, the ACE is identified from associational estimates by ensuring the absence of confounding. This is accomplished by statistically conditioning on a *sufficient adjustment set* of variables ($X$). Consequently, the ACE is expressed as $\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].

@@

<!-- 8. But which approach are we going to use? -->
The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of the ACE through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model*. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.

@@

<!-- 3. What are the advantages of DAGs? -->
DAGs offer two key advantages for modeling causal structures. Firstly, they represent causal relations in a nonparametric and fully interactive manner. This feature allows for feasible causal analysis strategies without needing the specification of the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with each building block individually [@McElreath_2020]. These building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].

@@

<!-- 5. continuing with the motivating example -->
The motivating example in @sec-framework-structural can further illustrate how to use the five fundamental building blocks to construct a system's causal structure. In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$. However, unlike the previous case, an experiment cannot be conducted, and the problem suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$. Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}. @fig-example illustrates the plausible causal structure of the motivating example. A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks. The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.

>>>>

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T=\{1,2\})$. Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding. The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness in comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006]. The Z-test is expressed as a signal-to-noise ratio: $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is the difference between the group sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y| T=1]$ and $E[Y| T=2]$, respectively. The noise $\hat{s}_{p}$ is the unpooled sample variability observed between the two groups.

@@

<!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
Experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. However, researchers often face constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these situations, causal inference provides a valuable alternative for generating causal estimates, particularly when the goal is to understand the mechanisms underlying specific data. Moreover, the framework provides significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].

<!-- 5. What is causal inference? -->
Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, *causal inference* is a framework designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. The framework employs rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009]. This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals, essential for understanding and defining causal effects. *Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a specific cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].

<!-- 6. But how does it work? --> 
Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality can be provided. The framework begins by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$. Notice when a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and thus termed a *counterfactual*. To overcome the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE). The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, representing the difference between the average of observed potential outcomes and counterfactuals across the sample. Finally, similar to experimental studies, the counterfactuals are identified from associational estimates by ensuring the absence of confounding. This is accomplished by statistically conditioning on a *sufficient adjustment set* of variables ($X$). As a result, the ACE is expressed as $\tau = E_{X}\left[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]\right]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].

@@

<!-- 8. But which approach are we going to use? -->
The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of a system through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. DAGs function as heuristics, effectively conveying the presumed causal structure of the system underlying the ACE, referred to as a *scientific model*. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.

@@

<!-- 3. What are the advantages of DAGs? -->
DAGs offer two key advantages for modeling causal structures. Firstly, they represent causal relations in a nonparametric and fully interactive manner. This feature allows for feasible causal analysis strategies without needing the specification of the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with individual building blocks [@McElreath_2020]. These building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].

@@

<!-- 5. continuing with the motivating example -->
The motivating example in @sec-framework-structural can be used to further illustrate how the five fundamental building blocks help construct a system's causal structure. In this scenario, an experiment cannot be conducted but the investigator still aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$. The problem further suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$. Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}. @fig-example illustrates the plausible causal structure of this motivating example. A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks. The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. Additionally, the figures feature multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
